\section{Getting Started}

\subsection{Prerequisites}

Before installing \Idris{}, you will need to make sure you have all of the necessary libraries and tools.
You will need:

\begin{itemize}
\item A fairly recent Haskell platform. Version 2013.2.0.0
should be sufficiently recent, though it is better to be completely up to
date.
\item The \emph{GNU Multiple Precision Arithmetic Library} (GMP) is available from MacPorts and all major Linux distributions.
\end{itemize}

\subsection{Downloading and Installing}

The easiest way to install \Idris{}, if you have all of the prerequisites, is to type:

\begin{lstlisting}[style=stdout]
cabal update; cabal install idris
\end{lstlisting}

\noindent
This will install the latest version released on Hackage, along with any dependencies.
If, however, you would like the most up to date development version you can find it, as well as build intructions, on GitHub at: \url{https://github.com/edwinb/Idris-dev}.

To check that installation has succeeded, and to write your first \Idris{} program, create a file called ``\texttt{hello.idr}'' containing the following text:

\inputIdrisListing{./examples/hello.idr}

\noindent
If you are familiar with Haskell, it should be fairly clear what the program is doing and how it works, but if not, we will explain the details later.
You can compile the program to an executable by entering \texttt{idris hello.idr -o hello} at the shell prompt.
This will create an executable called \texttt{hello}, which you can run:

\begin{lstlisting}[style=stdout]
$ idris hello.idr -o hello
$ ./hello
Hello world
\end{lstlisting}

\noindent
Note that the \texttt{\$} indicates the shell prompt!
Should the \Idris{} executable not be found please ensure that you have added \verb!~/.cabal/bin! to your \verb!$PATH! environment variable.
Mac OS X users may find they need to use \verb!~/Library/Haskell/bin! instead.
Some useful options to the \Idris{} command are:

\begin{itemize}
\item \verb!-o prog! to compile to an executable called \texttt{prog}.
\item \verb!--check! type check the file and its dependencies without starting the
interactive environment.
\item \verb!--help! display usage summary and command line options
\end{itemize}

\subsection{The Interactive Environment}

Entering \texttt{idris} at the shell prompt starts up the interactive environment.
You should see something like the following:

\begin{lstlisting}[style=stdout]
$ idris
     ____    __     _
    /  _/___/ /____(_)____
    / // __  / ___/ / ___/     Version ^\version{}^
  _/ // /_/ / /  / (__  )      http://www.idris-lang.org/
 /___/\__,_/_/  /_/____/       Type :? for help

Idris>
\end{lstlisting}

\noindent
This gives a \texttt{ghci}-style interface which allows evaluation of, as well as type checking of, expressions; theorem proving, compilation; editing; and various other operations.
The command \texttt{:?} gives a list of supported commands.
Listing~\ref{run1} shows an example run in which \texttt{hello.idr} is loaded, the type of \texttt{main} is checked and then the program is compiled to the executable \texttt{hello}.
Type checking a file, if successful, creates a bytecode version of the file (in this case \texttt{hello.ibc}) to speed up loading in future.
The bytecode is regenerated if the source file changes.

\begin{lstlisting}[caption={Sample Interactive Run}, label=run1, style=stdout, float=htp]
$ idris hello.idr
     ____    __     _
    /  _/___/ /____(_)____
    / // __  / ___/ / ___/     Version ^\version{}^
  _/ // /_/ / /  / (__  )      http://www.idris-lang.org/
 /___/\__,_/_/  /_/____/       Type :? for help

Type checking ./hello.idr
*hello> :t main
Main.main : IO ()
*hello> :c hello
*hello> :q
Bye bye
$ ./hello
Hello world
\end{lstlisting}

